---
title: "Untitled"
author: "Alec Pixton"
date: "2024-03-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(ggplot2)
library(MASS)
Data <- read.csv('diamonds4.csv', header = TRUE)

```

```{r}
Data$carat_cat <- ifelse(Data$carat < .4999 | Data$carat > 0.75 & Data$carat < 0.999 | Data$carat > 1.25 & Data$carat < 1.499 | Data$carat > 1.75 & Data$carat < 1.999 | Data$carat > 2.25 & Data$carat < 2.499 | Data$carat > 2.75 & Data$carat < 2.999 | Data$carat > 3.25 & Data$carat < 3.499 | Data$carat > 3.75 & Data$carat < 3.999 | Data$carat > 4.25 & Data$carat < 4.499 | Data$carat > 4.75 & Data$carat < 4.999 | Data$carat > 5.25 & Data$carat < 5.499 | Data$carat > 5.75 & Data$carat < 5.999 | Data$carat > 6.25 & Data$carat < 6.499 | Data$carat > 6.75 & Data$carat < 6.999 | Data$carat > 7.25 & Data$carat < 7.499, "low", "high")
```

```{r}
ggplot(Data, aes(x=carat,y=price))+
  geom_point()+
  geom_smooth(method='lm')
  labs(x="Carat", y="Price", title="Price by Carat Weight")
```

```{r}
result <- lm(price~carat, data=Data)
par(mfrow=c(2,2))
plot(result)
```

The variance of the residuals is not constant. The variance increases at higher values.
The residuals are not spread evenly across 0.
To stabilize the variance first we transform the response variable.

```{r}
MASS::boxcox(result,lambda=seq(.28,0.35,by=0.01))
```

Choose lambda = 0.31.

```{r}
Data$transformed_price = Data$price^0.31
```

```{r}
ggplot(Data, aes(x=carat,y=transformed_price))+
  geom_point()+
  geom_smooth(method='lm')
  labs(x="Carat", y="Price", title="Price by Carat Weight")
```

```{r}
price_transformed_result <- lm(transformed_price~carat, data=Data)
par(mfrow=c(2,2))
plot(price_transformed_result)
```

```{r}
Data$log_trans_price <- log(Data$price)
```

```{r}
ggplot(Data, aes(x=carat,y=log_trans_price))+
  geom_point()+
  geom_smooth(method='lm')
  labs(x="Carat", y="Price", title="Price by Carat Weight")
```


```{r}
log_transformed_result <- lm(log_trans_price~carat, data=Data)
par(mfrow=c(2,2))
plot(log_transformed_result)
```

```{r}
Data$log_carat <- log(Data$carat)
```

```{r}
log_log_tranform <- lm(log_trans_price~log_carat, data = Data)
par(mfrow=c(2,2))
plot(log_log_tranform)
```

```{r}
summary(log_log_tranform)
```


```{r}
ggplot(Data, aes(x=log_carat,y=log_trans_price))+
  geom_point()+
  geom_smooth(method='lm')
  labs(x="Carat", y="Price", title="Price by Carat Weight")
```



**carat_cat**

```{r}
high_carat <- which(Data$carat_cat=='high')
```






```{r}
ggplot(Data, aes(x = log_carat, y = log_trans_price)) +
  geom_point(aes(color = carat_cat), alpha = 0.5) +
  theme_minimal() +
  labs(title = "Carat vs. Price by Color", x = "Carat", y = "Price")
```









```{r}
high_carat <- lm(log_trans_price~log_carat, data = Data, subset = carat_cat == 'high')
par(mfrow=c(2,2))
plot(log_log_tranform)
```

```{r}
summary(high_carat)
```

```{r}
low_carat <- lm(log_trans_price~log_carat, data = Data, subset = carat_cat == 'low')
par(mfrow=c(2,2))
plot(log_log_tranform)
```

```{r}
summary(low_carat)
```

```{r}
confint(log_log_tranform, param = 'log_carat', level=0.95)
```
Both the low and high carat 

h0: B1 = 1.944020
ha: B1 != 1.944020

```{r}
(1.90820 - 1.944020)/0.02104
```

```{r}
qt(.95, df = 528)
```

The t-statistic is greater than the critical value so we reject the null hypothesis, there is evidence to support 'low' diamonds being more expensive than the average.

h0: B1 = 1.944020
ha: B1 != 1.944020

```{r}
(2.00343 - 1.944020)/0.01851
```

The t-statistic is greater than the critical value so we reject the null hypothesis, there is evidence to support 'high' diamonds being more expensive than the average.