---
title: "Untitled"
author: "Darreion Bailey"
date: "2024-03-22"
output: html_document
---

```{r, echo=FALSE, message=FALSE, warning=FALSE, results='hide'}
library(tidyverse)
library(gridExtra)
library(MASS)
df <- read.csv("dataset.csv")
df$cut <- factor(df$cut, levels = c("Astor Ideal", "Ideal", "Very Good", "Good"))
df$clarity<-factor(df$clarity, levels = c("FL", "IF", "VVS1", "VVS2", "VS1", "VS2", "SI1", "SI2"))
```

```{r}
library(ggplot2)
```

```{r}
df$carat_cat <- ifelse(df$carat < .4999 | df$carat > 0.75 & df$carat < 0.999 | df$carat > 1.25 & df$carat < 1.499 | df$carat > 1.75 & df$carat < 1.999 | df$carat > 2.25 & df$carat < 2.499 | df$carat > 2.75 & df$carat < 2.999 | df$carat > 3.25 & df$carat < 3.499 | df$carat > 3.75 & df$carat < 3.999 | df$carat > 4.25 & df$carat < 4.499 | df$carat > 4.75 & df$carat < 4.999 | df$carat > 5.25 & df$carat < 5.499 | df$carat > 5.75 & df$carat < 5.999 | df$carat > 6.25 & df$carat < 6.499 | df$carat > 6.75 & df$carat < 6.999 | df$carat > 7.25 & df$carat < 7.499, "low", "high")
```

```{r, echo=FALSE, message=FALSE, warning=FALSE, results='hide'}
df$log_trans_price <- log(df$price)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE, results='hide'}
df$log_carat <- log(df$carat)
```

# Visualizations

## 1. Data & Variables

This dataset provided by Blue Nile describes more than 1200 different diamonds that are for sale on their website, and is a reflection in some ways of their standards as a diamond merchant. The dataset is in a .csv format. The following are the variables which comprise the dataset: Carat, cut, clarity, color, and price. An additional variable (carat_cat) was also created to address a claim in the next section.

Carat refers to a diamond's weight, which is a quantitative variable. Cut refers to the quality of the dimensions of a diamond (i.e. how the surfaces are positioned to create sparkles) and consists of 4 grade categories: Astor Ideal, Ideal, Very good, and Good. Clarity refers to the number of imperfections internally and externally about or within the stone itself, and consists of 11 grades, 8 of which Blue Nile sells: SI1, SI2, VS1, VS2, VVS1, VVS2, IF, and FL diamonds, with FL diamonds being the most rare and expensive. Color refers to the purity of the color, and Blue Nile's business consists of 3 categories across 8 grades: Colorless diamonds (D, E, and F grades), Near-colorless diamonds (G, H, I, and J), and Faint color diamonds (K).

## 2. Price vs. 4Cs & Addressing Claims 

### Cut

Using *Figure 1* (*Cut vs. Price*), we can see that Astor Ideal cuts have a higher average price than any other cut, which is to be expected given that it is the highest cut grade. We can also see that Ideal and Very good are almost identical, with Ideal having a slightly higher average price than Very Good. From this, we gather that a higher cut grade does correlate to a higher price linearly.

Using *Figure 4 (Carat vs. Price)* - where cut is added for context - we can see that the Ideal cut dominates the upper half of the distribution. This fact, coupled with the tightness of the distribution and the steep gradient (density) seen in *Figure 1 (Cut vs. Price)*, confirms the claim on the Blue Nile website that the cut of the diamond **can be** the biggest factor on price.

### Clarity

Using *Figure 2* (*Clarity vs. Price*), we can see that Flawless (FL) diamond clarities are the most expensive. Given that there are so few FL clarity diamonds in this dataset, this would also give credence to the claim made by Blue Nile that less than 1% of all diamonds are FL clarity, though this is hardly provable given our sample size.

Additionally, using *Figure 5 (Frequency of Diamond Clarity Grades)*, we can see that the distribution of diamond clarities are such that VS1 and VS2 are the most popular clarity among Blue Nile, which confirms their claim as such.

### Color

Using *Figure 3 (Color vs. Price)*, we can see that D, E, and F color grades are by far the most expensive, which is to be expected given that they are the highest quality color grades according to Blue Nile. Additionally, we see that G, H, and I price values hover around the midpoint, which confirms their claim that these color grades are a great value for their quality.

However, as we can see in *Figure 6 (Frequency of Diamond Color Grades)*, Blue Nile appears to sell the least amount of J color grade diamonds, which fails to confirm their claim that D, E, and F color grades are the rarest among all color grades. Although, this dataset is not a representation of all diamonds harvested, and this data could be a reflection of Blue Nile standards and marketing influence.

### Carat

Using *Figure 4 (Carat vs. Price) -* where cut is added for context - we see a very tight distribution and a steep gradient. This indicates that carat has a very strong relationship with price. In comparison to the other of the 4Cs, the distributions are not quite as tight, and the gradient is not quite as steep, confirming their claim that carat weight is the biggest factor on price.

Using *Figure 7 (Carat vs. Price by Carat_cat)*, we can infer based on the visualization that given the density of data points around half and full carat increments that the claim on the Blue Nile website which suggests that buying slightly below half-carat and whole-carat values will save significant money for little compromise is confirmed. This point will be further addressed in detail in SLR section of this report.

### Hierarchy

\<generate hierarchy of all 4 vs. price\>

We can also see from these comparisons that cut is the second biggest factor on price overall, behind carat weight.

```{r}
# Box plot for Cut vs. Price
fig1<-ggplot(df, aes(x = cut, y = price)) +
  geom_violin(aes(fill = cut), show.legend = FALSE) +
  theme_minimal() +
  theme(plot.caption = element_text(hjust=0.5)) +
  labs(title = "Cut vs. Price", caption="Figure 1", x = "Cut", y = "Price")

```

```{r}
# Box plot for Clarity vs. Price
fig2<-ggplot(df, aes(x = clarity, y = price)) +
  geom_violin(aes(fill = clarity), show.legend = FALSE) +
  theme_minimal() +
  theme(plot.caption = element_text(hjust=0.5)) +
  labs(title = "Clarity vs. Price", caption ="Figure 2", x = "Clarity", y = "Price")

```

```{r}
# Box plot for Color vs. Price
fig3<-ggplot(df, aes(x = color, y = price)) +
  geom_violin(aes(fill = color), show.legend = FALSE) +
  theme_minimal() +
  theme(plot.caption = element_text(hjust=0.5)) +
  labs(title = "Color vs. Price", caption="Figure 3", x = "Color", y = "Price")

```

```{r}
# Scatter plot for Carat vs. Price
fig4<-ggplot(df, aes(x = carat, y = price)) +
  geom_point(aes(color = cut), alpha = 0.5) + # Color points by cut to add more context
  theme_minimal() +
  theme(plot.caption = element_text(hjust=0.5)) +
  labs(title = "Carat vs. Price", caption="Figure 4: 'Cut' added for more context", x = "Carat", y = "Price")

```

```{r}
grid.arrange(fig1,fig2,fig3,fig4, ncol=2)
```

```{r}
# Create a bar chart for the clarity variable
ggplot(df, aes(x = clarity)) +
  geom_bar(fill = "orange", color = "black") +
  theme_minimal() +
  theme(plot.caption = element_text(hjust=0.5)) +
  labs(title = "Frequency of Diamond Clarity Grades", caption="Figure 5", x = "Clarity", y = "Frequency")

```

```{r}
# Create a bar chart for the color variable
ggplot(df, aes(x = color)) +
  geom_bar(fill = "green", color = "black") +
  theme_minimal() +
  theme(plot.caption = element_text(hjust=0.5)) +
  labs(title = "Frequency of Diamond Color Grades", caption="Figure 6", x = "Color", y = "Frequency")

```

```{r}
ggplot(Data, aes(x = log_carat, y = log_trans_price)) +
  geom_point(aes(color = carat_cat), alpha = 0.5) +
  theme_minimal() +
  theme(plot.caption = element_text(hjust=0.5)) +
  labs(title = "Carat vs. Price by Carat_cat", caption="Figure 7", x = "Carat", y = "Price")
```
